Ornaments as indicators of social changes in northeastern Taiwan before and after the European colonial period

### 2020-04-23 analysis/paper/paper.Rmd
correct the AD age

This article describes personal ornaments excavated from the upper component of Kiwulan (1400 AD-1900 AD), the largest Iron Age settlement on the Yilan plain in northeastern Taiwan. The first recorded European presence in Yilan was a Spanish revenge attack on Indigenous villages in 1632 [@Borao2001, pp. 163]. In 1647 the Dutch attacked villages and forced them to accept colonial rules and pay an annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest Indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. Following defeat of Dutch by the Chinese general Koxinga in 1661-1662, the Dutch abandoned northern Taiwan. Direct contact with Han Chinese is indicated by Qing dynasty census reports mentioning Yilan villages in 1821 [@Yao1996].

@@

The distribution of frequencies for subtypes in each major class is presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of the effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bells and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells have less variety typically >2 cm long with a wide variety of human faces as a motif. Although glass beads have less variety in size, small (0.5-1 cm) and medium (1-2 cm), they have a wide variety of colors or patterns mostly made by a winding technique and with high lead content indicating possibly from China [@Cheng2008]. Although we are not certain of the specific origin of these beads, research suggest that these glass beads and metal ornaments have similar production techniques and composition to those found in China [@Chen2011; @Wang2018]. There seem to be no obvious changes in the sources of glass beads or metal ornaments at different periods in the upper component of Kiwulan (1400-1900 AD). However, the glass beads found from the lower component (700-1200 AD) are mostly Indo-Pacific beads, widespread in Southeast Asian sites from 300 BC until the early 2nd millennium [@Francis2002; @Wang2018].

>>>>

This article describes personal ornaments excavated from the upper component of Kiwulan (1350 AD-1850 AD), the largest Iron Age settlement on the Yilan plain in northeastern Taiwan. The first recorded European presence in Yilan was a Spanish revenge attack on Indigenous villages in 1632 [@Borao2001, pp. 163]. In 1647 the Dutch attacked villages and forced them to accept colonial rules and pay an annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest Indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. Following defeat of Dutch by the Chinese general Koxinga in 1661-1662, the Dutch abandoned northern Taiwan. Direct contact with Han Chinese is indicated by Qing dynasty census reports mentioning Yilan villages in 1821 [@Yao1996].

@@

The distribution of frequencies for subtypes in each major class is presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of the effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bells and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells have less variety typically >2 cm long with a wide variety of human faces as a motif. Although glass beads have less variety in size, small (0.5-1 cm) and medium (1-2 cm), they have a wide variety of colors or patterns mostly made by a winding technique and with high lead content indicating possibly from China [@Cheng2008]. Although we are not certain of the specific origin of these beads, research suggest that these glass beads and metal ornaments have similar production techniques and composition to those found in China [@Chen2011; @Wang2018]. There seem to be no obvious changes in the sources of glass beads or metal ornaments at different periods in the upper component of Kiwulan (1350-1850 AD). However, the glass beads found from the lower component (650-1150 AD) are mostly Indo-Pacific beads, widespread in Southeast Asian sites from 300 BC until the early 2nd millennium [@Francis2002; @Wang2018].
